Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
Next figure.

There are two major components in PyPy:

\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}

In particular, we have defined a subset of the Python language called
"restricted Python" or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

@@

  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
             then a stack of transformations
  ]
\end{verbatim}

\begin{enumerate}
\item We take as input RPython functions\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This allows us to use unrestricted
       python for meta-programming purposes at load time, in a
       seemingly staged programming approach, in which the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.}, and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

\item We perform type inference on the control flow graphs.  At this
      stage, types inferred are part of the type system which is the
      very definition of the RPython sub-language: they are roughly a
      subset of Python's built-in types, with some more precision to
      describe e.g. the items stored in container types.
      Occasionally, a single input function can produce several
      specialized versions, i.e. several similar but differently typed
      graphs.  This type inference process is described in more
      details in section \ref{typeinference}.
\end{enumerate}

At the end of the front-end analysis, the input RPython program is
represented as a forest of flow graphs with typed variables.  Following
this analysis are a number of transformation steps.  Each transformation
step modifies the graphs in-place, by altering their structure and/or
the operations they contain.  Each step inputs graphs typed in one type
system and leaves them typed in a possibly different type system, as we
will describe in the sequel.  Finally, a back-end turns the resulting
graphs into code suitable for the target environment, e.g. C source code
ready to be compiled.

@@

For example, \texttt{ll\textunderscore{}append()} in figure \ref{llappend}
is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

\begin{figure}
\begin{verbatim}
def ll_append(lst, newitem):
  # Append an item to the end of the vector.
  index = lst.length       # get the 'length' field
  ll_resize(lst, index+1)  # call another helper
  itemsarray = lst.items   # get the 'items' field
  itemsarray[index] = item # behaves like a C array
\end{verbatim}
\caption{A helper to implement \texttt{list.append()}.}
\label{llappend}
\end{figure}

@@

The four levels that we considered so far are summarized in figure 2.

    [figure 2:    RPython
                  /     \
                 /       \
       LLTypeSystem     OOTypeSystem
               /
              /
      Raw addresses
    ]
\end{verbatim}

@@

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:

\begin{enumerate}
\item the types, which we use to tag the variables of the graphs at
      the given level.  (Types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly.)

\item the Python objects that emulate instances of these types.  (More
      about them below.)
\end{enumerate}

We have defined well-typed operations between variables of these types,
plugging on the standard Python operators.  These operations are the
ones that the emulating instances implement.  As seen above, the types
can also be used by type inference when analysing system code like the
helpers of figure 1.

@@

The set $A$ of RPython types is defined as the following formal terms:

\begin{itemize}
\item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
      to "impossible value" and "most general value");

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}

Figures 6 and 7 shows how these types are ordered to form a lattice.  We
mostly use its structure of [join-semilattice] only.

>>>>

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
"restricted Python" or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

@@

  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
             then a stack of transformations
  ]
\end{verbatim}
%
\begin{enumerate}
\item We take as input RPython functions\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This allows us to use unrestricted
       python for meta-programming purposes at load time, in a
       seemingly staged programming approach, in which the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.}, and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

\item We perform type inference on the control flow graphs.  At this
      stage, types inferred are part of the type system which is the
      very definition of the RPython sub-language: they are roughly a
      subset of Python's built-in types, with some more precision to
      describe e.g. the items stored in container types.
      Occasionally, a single input function can produce several
      specialized versions, i.e. several similar but differently typed
      graphs.  This type inference process is described in more
      details in section \ref{typeinference}.
\end{enumerate}
 %     
At the end of the front-end analysis, the input RPython program is
represented as a forest of flow graphs with typed variables.  Following
this analysis are a number of transformation steps.  Each transformation
step modifies the graphs in-place, by altering their structure and/or
the operations they contain.  Each step inputs graphs typed in one type
system and leaves them typed in a possibly different type system, as we
will describe in the sequel.  Finally, a back-end turns the resulting
graphs into code suitable for the target environment, e.g. C source code
ready to be compiled.

@@

For example, \texttt{ll\textunderscore{}append()} in figure \ref{fig_llappend}
is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

\begin{figure}
\begin{verbatim}
def ll_append(lst, newitem):
  # Append an item to the end of the vector.
  index = lst.length       # get the 'length' field
  ll_resize(lst, index+1)  # call another helper
  itemsarray = lst.items   # get the 'items' field
  itemsarray[index] = item # behaves like a C array
\end{verbatim}
\caption{A helper to implement \texttt{list.append()}.}
\label{fig_llappend}
\end{figure}

@@

The four levels that we considered so far are summarized in figure
\ref{fig_typesystems}.

\begin{figure}
\begin{verbatim}
                      RPython
                      /     \
           LLTypeSystem     OOTypeSystem
                  /
          Raw addresses
\end{verbatim}
\caption{Type systems, informally ordered by abstraction level}
\label{fig_typesystems}
\end{figure}

@@

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:
%
\begin{enumerate}
\item the types, which we use to tag the variables of the graphs at
      the given level.  (Types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly.)

\item the Python objects that emulate instances of these types.  (More
      about them below.)
\end{enumerate}
%
We have defined well-typed operations between variables of these types,
plugging on the standard Python operators.  These operations are the
ones that the emulating instances implement.  As seen above, the types
can also be used by type inference when analysing system code like the
helpers of figure \ref{fig_llappend}.

@@

The set $A$ of RPython types is defined as the following formal terms:
%
\begin{itemize}
\item $Bot$, $Top$ -- the minimum and maximum elements (corresponding
      to "impossible value" and "most general value");

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}
%
Figures 6 and 7 shows how these types are ordered to form a lattice.  We
mostly use its structure of [join-semilattice] only.
